Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-04-11 VegOzone.tex
Update plots

\begin{abstract}
  The unique vegetation of the subarctic region acclimatized to extremes of cold and midnight sun are likely to be at threat from the combined impacts of climate change and increasing ozone concentrations \chem{[O_3]}. The atmospheric and climatic characteristics of the subarctic are known to lead to pronounced peak \chem{[O_3]} in spring. To date, only a few studies assessed the response of subarctic vegetation to variations in climate and air pollution. This study looks to fill this knowledge gap by examining the uncertainty in risk assessments that are using more general boreal paramterizations. With no bespoke parameterization of stomatal response of subarctic vegetation to essential climate variables available, we propose an adjustment of existing, Boreal parameterizations, based on local climate condition. To this end, we vary temperature response functions according to the climatological temperature distribution within the growing season at our site. We evaluate the extent to which two recent years (2018 and 2019) deviate from climatic and \chem{[O_3]} norms and how these potentially more frequent future deviations may influence the projected ozone damage to subarctic vegetation. We find that 2018 was an anomalously warm and bright year, particularly in spring and early summer. Higher than average \chem{[O_3]} was observed in April/May while frequent episodes of ozone volume mixing ratios (VMRs) above $40\,\unit{ppb}$ occurred in June--August. These episodes are in part attributable to forest fires in the Northern Hemisphere and warmer and sunnier conditions. We apply the integrated flux-metric Phytotoxic Ozone Dose (POD) to determine ozone risk and damage to vegetation as a function of \chem{[O_3]}, environmental factors, and species-specific physiology. Our study suggests that using standard boreal parameterizations in assessments likely leads to underestimating the risk of ozone damage in this region. We find that local climate adjusted parameterizations of plant functional types (PFTs) for subarctic vegetation result in an ozone-induced biomass loss of $2.5$ to $17.4\,\unit{\%}$. For some species, this loss is up to $xx\,\unit{\%}$ larger than projected from standard boreal parameterizations used in general assessments. Efforts should be targeted towards accurately defining subarctic vegetations' physiological response to essential climate variables.
\end{abstract}

@@

To fill gaps in the ozone data, we applied a methodology based on a Reynolds decomposition of the form:
\begin{equation}
  \chem{[O_3]} = \chem{\left<[O_3]\right>} + \chem{\Delta[O_3]},
  \label{eq:reynolds_decomp}
\end{equation}
with an ozone climatology \chem{\left<[O_3]\right>} and ozone anomalies (deviation from the norm) \chem{\Delta[O_3]}. We computed a generalized climatology for northern Fennoscandia from long-term records of different close by sites and for our site. We determined the station with the highest correlation coefficient compared to Svanhovd. For this site, we computed the anomalies with respect to the generalized climatology for the gap period. These anomalies were than added to the climatology at out site and corrected for an increase in present-day tropospheric background ozone compared to the early 1990s. For more details refer to \citet{ACPD:Falk2021}.

@@

As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a, \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition only started in late April. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak \chem{[O_3]}. A method for gap-filling these data has been presented in \citet{ACPD:Falk2021}. In contrast, \chem{[O_3]} only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{data_svanvik_2018_2019}
  %\includegraphics[width=12cm]{fig02}
  \caption{Observational data from atmospheric monitoring at Svanhovd in 2018/19. The hatched areas indicate periods without ozone monitoring data. (a) Hourly averaged \chem{[O_3]}; (b) hourly averaged temperature; (c) daily accumulated precipitation; (d) hourly averaged global irradiance.}
  \label{fig:data_svanvik_2018_2019}
\end{figure*}

@@

\begin{figure}[t]
  \includegraphics[width=12cm, clip, trim=1cm 0 1cm 1cm]{climatediagram_box}
  %\includegraphics[width=8.3cm]{fig04}
  \caption{Climate diagram. The diagram is based on climatological data for Svanvik/Pasvik (1992--2012). (a) Monthly temperatures. (b) Monthly accumulated precipitation.}
  \label{fig:climatediagram}
\end{figure}

@@

%Ozone
The daily mean ozone climatology (\chem{\left<[O_3]\right>}) (1986--1996) is shown in Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}. Darker colors indicate higher probability to observe these values. On top of the density distributions, a $10\,\unit{days}$ average of daily mean (\chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}}) is displayed together with $1 \sigma$ uncertainties and SE, respectively. A spline was fitted through the data to guide the eye. The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display the expected, pronounced seasonal cycle. The spring peak occurs on average on day of the year (\unit{doy}) $100$ and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $28\,\unit{ppb}$. The decline in \chem{\left<[O_3]\right>} coincides with the average beginning of \chem{CO_2} uptake by coniferous trees \citep{TB:Kolari2007, TP:Wallin2013}. In July--September (\unit{doy}~182--273), ozone is occasionally almost completely depleted. This hints to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{ozone_climatology_svanvik_obs_spline}
  %\includegraphics[width=8.3cm]{fig05}
  \caption{Svanhovd ozone climatology. The density distribution is shown together with a multi-annual average of daily \chem{[O_3]} (climatology). A spline has been fitted through daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

\subsection{Derivation from climate norm in the 2018/19 growing season}
\label{subsec:anomalies}
In the following, we look at the 2018/19 anomalies and discuss the differences between the two years. As weather extremes like in 2018 are likely to become more frequent with climate change, this will help to assess possible future changes in the vulnerability of subarctic vegetation to ozone.\\

@@

\begin{figure}[t]
  %\includegraphics[width=12cm]{fig06}
  \centering
  (a)\hspace{8.3cm}(b)\\
  \includegraphics[width=8.3cm]{temperature_signific_1.pdf}
  \includegraphics[width=8.3cm]{precipitation_signific_1.pdf}\\
  (c)\hspace{8.3cm}(d)\\
  \includegraphics[width=8.3cm]{global_rad_signific_1.pdf}
  \includegraphics[width=8.3cm]{ozone_signific_1.pdf}
  \caption{2018/19 anomalies of key environmental variables at Svanhovd displayed as percentages of days significantly deviating from climatological mean through standard deviation for each month. Negative deviations from the climatology are shown as a negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The hatched area between the dashed lines indicates statistical significance under the assumption of a normal distribution ($15.9\,\unit{\%}$, $30.8\,\unit{\%}$ for Precip.). (a) Temperature; (b) precipitation; (c) global irradiance; (d) ozone.}
  \label{fig:anomalies_svanvik}
\end{figure}

@@

%Ozone
We use the bias-corrected and cross-calibrated ozone climatology \citep{ACPD:Falk2021} and assess the monthly significance of the ozone concentration anomalies in 2018/19. The resulting percentage of days above/below $\pm 1\,\sigma$ is displayed for each month in Fig.~\ref{fig:anomalies_svanvik}d. Note that missing bars for months in winter and spring are both due to the temporal extent of our data and anomalies not meeting the $1\sigma$ criteria. The star indicates the reconstructed data in July. In April/May 2018, ozone was significantly elevated, while no significant enhancement is found in 2019. In both years, ozone concentrations were significantly lower than average in November. These results suggest that stagnant meteorological conditions in April/May in combination with a late start of the growing season contributed more to elevated ozone than the extensive forest fires in Sweden that did not significantly enhance \chem{[O_3]} in 2018.

@@

\section{Modeling results and implications on ozone risk assessments}
\label{sec:do3se_results}

>>>>

\begin{abstract}
  The unique vegetation of the subarctic region acclimatized to extremes of cold and midnight sun are likely to be at threat from the combined impacts of climate change and increasing ozone concentrations \chem{[O_3]}. The atmospheric and climatic characteristics of the subarctic are known to lead to pronounced peak \chem{[O_3]} in spring. To date, only a few studies assessed the response of subarctic vegetation to variations in climate and air pollution. This study looks to fill this knowledge gap by examining the uncertainty in risk assessments that are using more general boreal paramterizations. With no bespoke parameterization of stomatal response of subarctic vegetation to essential climate variables available, we propose an adjustment of existing, Boreal parameterizations, based on local climate condition. To this end, we vary temperature response functions according to the climatological temperature distribution within the growing season at our site. We evaluate the extent to which two recent years (2018 and 2019) deviate from climatic and \chem{[O_3]} norms and how these potentially more frequent future deviations may influence the projected ozone damage to subarctic vegetation. We find that 2018 was an anomalously warm and less cloudy year, particularly in spring and early summer. Higher than average \chem{[O_3]} was observed in April/May while frequent episodes of ozone volume mixing ratios (VMRs) above $40\,\unit{ppb}$ occurred in June--August. These episodes are in part attributable to forest fires in the Northern Hemisphere and warmer and sunnier conditions. We apply the integrated flux-metric Phytotoxic Ozone Dose (POD) to determine ozone risk and damage to vegetation as a function of \chem{[O_3]}, environmental factors, and species-specific physiology. Our study suggests that using standard boreal parameterizations in assessments likely leads to underestimating the risk of ozone damage in this region. We find that local climate adjusted parameterizations of plant functional types (PFTs) for subarctic vegetation result in an ozone-induced biomass loss of $2.5$ to $17.4\,\unit{\%}$. For some species, this loss is up to $xx\,\unit{\%}$ larger than projected from standard boreal parameterizations used in general assessments. Efforts should be targeted towards accurately defining subarctic vegetations' physiological response to essential climate variables.
\end{abstract}

@@

To fill gaps in the ozone data, we applied a methodology based on a Reynolds decomposition of the form:
\begin{equation}
  \chem{[O_3]} = \chem{\left<[O_3]\right>} + \chem{\Delta[O_3]},
  \label{eq:reynolds_decomp}
\end{equation}
with an ozone climatology \chem{\left<[O_3]\right>} and ozone anomalies (deviation from the norm) \chem{\Delta[O_3]}. We computed a generalized climatology for northern Fennoscandia from long-term records of different close by sites and for our site. We determined the station with the highest correlation coefficient compared to Svanhovd. For this site, we computed the anomalies with respect to the generalized climatology for the gap period. These anomalies were than added to the climatology at out site and corrected for an increase in present-day tropospheric background ozone compared to the early 1990s. For more details refer to \citet{ACP:Falk2021}.

@@

As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a, \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, since data acquisition only started in late April. In summer 2018 (June--August), high ozone concentrations ($\chem{[O_3]} > 40\,\unit{ppb}$) were recorded 50 times. The highest summer ozone concentration ($\chem{[O_3]} = 50.2\,\unit{ppb}$) was measured on July 25. This coincides with the period of the most extensive forest fires in central Sweden which occurred from July 12--29 \citep{SOU2019}. However, due to the above-mentioned data acquisition problems, we missed most of the corresponding ozone data for this event and most likely also the peak \chem{[O_3]}. A method for gap-filling these data has been presented in \citet{ACP:Falk2021}. In contrast, \chem{[O_3]} only rose 18 times above the threshold of $40\,\unit{ppb}$ during the summer of 2019.

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{data_svanvik_2018_2019}
  %\includegraphics[width=12cm]{fig02}.
  \caption{Observational data from atmospheric monitoring at Svanhovd in 2018/19. The hatched areas indicate periods without ozone monitoring data. (a) Hourly averaged ozone VMR; (b) hourly averaged temperature; (c) daily accumulated precipitation; (d) hourly averaged global irradiance.}
  \label{fig:data_svanvik_2018_2019}
\end{figure*}

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{climatediagram_box}
  %\includegraphics[width=8.3cm]{fig04}
  \caption{Climate diagram based on meteorological data for Svanvik/Pasvik (1992--2012). Ozone climatology based on ozone monitoring data at NIBIO Environment Centre Svanhovd(1986--2003). A box indicates the upper/lower quartile, whiskers the $1.5\times$ Interquartile Range (IQR), and circles outliers. The median is marked by a colored horizontal line and the mean by a triangle. Notches indicate the confidence interval around the median. (a) Monthly temperatures; (b) Monthly accumulated precipitation; (c) Monthly global radiation; (d) Monthly ozone VMR}
  \label{fig:climatediagram}
\end{figure*}

@@

%Ozone
The daily mean ozone climatology (\chem{\left<[O_3]\right>}) (1986--1996) is shown in Figure~\ref{fig:climatediagram}d. Darker colors indicate higher probability to observe these values. On top of the density distributions, a $10\,\unit{days}$ average of daily mean (\chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}}) is displayed together with $1 \sigma$ uncertainties and SE, respectively. A spline was fitted through the data to guide the eye. The \chem{[O_3]} density distribution and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} display the expected, pronounced seasonal cycle. The spring peak occurs on average on day of the year (\unit{doy}) $100$ and amounts to $40\,\unit{ppb}$ while the annual average \chem{\left<[O_3]\right>} is $28\,\unit{ppb}$. The decline in \chem{\left<[O_3]\right>} coincides with the average beginning of \chem{CO_2} uptake by coniferous trees \citep{TB:Kolari2007, TP:Wallin2013}. In July--September (\unit{doy}~182--273), ozone is occasionally almost completely depleted. This hints to ozone uptake by vegetation even at low light intensities during midnight sun conditions in combination with stable planetary boundary layer conditions preventing mixing of ozone rich air.

\subsection{2018/19 growing season anomalies}
\label{subsec:anomalies}
In the following, we look at the 2018/19 anomalies and discuss the differences between the two years. As weather extremes like in 2018 are likely to become more frequent with climate change, this will help to assess possible future changes in the vulnerability of subarctic vegetation to ozone.\\

@@

\begin{figure*}[t]
  %\includegraphics[width=12cm]{fig06}
  \centering
  (a)\hspace{6cm}(b)\\
  \includegraphics[width=6cm]{temperature_signific_1.pdf}
  \includegraphics[width=6cm]{precipitation_signific_1.pdf}\\
  (c)\hspace{6cm}(d)\\
  \includegraphics[width=6cm]{global_rad_signific_1.pdf}
  \includegraphics[width=6cm]{ozone_signific_1.pdf}
  \caption{2018/19 anomalies of key environmental variables at Svanhovd displayed as percentages of days significantly deviating from climatological mean through standard deviation for each month. Negative deviations from the climatology are shown as a negative percentage. The annual positive/negative deviations are indicated in the respective corners (right upper/right lower). The hatched area between the dashed lines indicates statistical significance under the assumption of a normal distribution ($15.9\,\unit{\%}$, $30.8\,\unit{\%}$ for Precip.). (a) Temperature; (b) precipitation; (c) global irradiance; (d) ozone.}
  \label{fig:anomalies_svanvik}
\end{figure*}

@@

%Ozone
We use the bias-corrected and cross-calibrated ozone climatology \citep{ACP:Falk2021} and assess the monthly significance of the ozone concentration anomalies in 2018/19. The resulting percentage of days above/below $\pm 1\,\sigma$ is displayed for each month in Fig.~\ref{fig:anomalies_svanvik}d. Note that missing bars for months in winter and spring are both due to the temporal extent of our data and anomalies not meeting the $1\sigma$ criteria. The star indicates the reconstructed data in July. In April/May 2018, ozone was significantly elevated, while no significant enhancement is found in 2019. In both years, ozone concentrations were significantly lower than average in November. These results suggest that stagnant meteorological conditions in April/May in combination with a late start of the growing season contributed more to elevated ozone than the extensive forest fires in Sweden that did not significantly enhance \chem{[O_3]} in 2018.

@@

\section{Results}
\label{sec:results}

\subsection{Parameter localization}
\label{subsec:res_local}

\section{Implications on ozone risk assessments}
\label{sec:do3se_results}
